---
title: "Project 1: Data Analysis"
author: "Puja Roy"
date: "`r Sys.Date()`"
output: openintro::lab_report
---

### Introduction
The goal of this project is to create an R Markdown file that generates a .CSV file containing all of the information for all of the chess players.

### Import the Libraries
```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
```

### Load the data
```{r}
# Read the text file
data <- read_lines("https://raw.githubusercontent.com/pujaroy280/DATA607Project1/main/tournamentinfo.txt", skip=4)
head(data)
```

### Clean & Transform the Data

I cleaned and transformed the data by deleting the divider lines and extracting the values into a dataframe. I also worked on categorizing the dataframe into even and odd rows and later merging the dataframes by each column.
```{r}
dividers <- which(grepl("^-", data))
data_transformed <- data[-dividers]
```

```{r}
data <- read.table(text = data_transformed, sep = "|", col.names = c("Pair", "Name", "Points", "R1", "R2", "R3", "R4", "R5", "R6", "R7", "NA"))
data <- data[1:(length(data) - 1)]
print(data)
```

```{r}
row_1 <- data %>% filter(row_number() %%2 == 1)
row_2 <- data %>% filter(row_number() %%2 == 0)
data_w <- cbind(row_1, row_2)
colnames(data_w) = c("Pair", "Name", "Points", "R1", "R2", "R3", "R4", "R5", "R6", "R7", "State", "ID_Rating", "Points_2", "R1_2", "R2_2", "R3_2", "R4_2", "R5_2", "R6_2", "R7_2")
print(data_w)
```

I leveraged Regex to create new columns of each round to identify Wins, Losses or Draws and likewise for the column pre_ratings.
```{r}
data_w <- data_w %>%
  separate_wider_regex(
    c(R1:R7),
    patterns = c(
      "[BDHLUWX] {2,4}",
      # Create new columns by extracting data using Regex expressions to identify Wins, Losses and Draws from Rounds 1-7.
      opponent_ID = "(?:\\d+)?"
    ),
     # Include new colums to identify which Rounds in the chess game had Wins, Losses or Draws
    names_sep = "_"
  )
print(data_w)
```

```{r}
data_w <- data_w %>%
  separate_wider_regex(
    ID_Rating,
    patterns = c(
      " \\d{8} / R: +",
      pre_ratings = "\\d+",
      "[ P0-9]*-> *",
      "\\d+[ P0-9]*"
    )
  )
print(data_w)
```

```{r}
# Replace missing values with NA
data_w <- replace(data_w, data_w=='', NA)
print(data_w)
```

```{r}
data_w$pre_ratings <- as.numeric(data_w$pre_ratings)
```

```{r}
data_w$Pair <- str_trim(data_w$Pair)
```

### Update the data by Calculating average pre-rating of opponents

Once the data was cleaned and transformed, I calculated the average pre-rating of each player’s opponents.
```{r}
pre_ratings_updated <- data_w$pre_ratings
names(pre_ratings_updated) <- data_w$Pair
print(pre_ratings_updated)
```

```{r}
data_w <- data_w %>%
  rowwise %>%  
  mutate(
      total_opponent_pre_ratings = sum(across(
            .cols = ends_with("_opponent_ID"),
            .fns = ~ unname(pre_ratings_updated[.x])),
          na.rm = TRUE),
      
      num_of_rounds = ncol(select(data_w, starts_with("R"))), # Count the number of rounds
      
      average_opponent_pre_rating = ifelse(num_of_rounds == 0, NA, 
                                           floor(total_opponent_pre_ratings / 7)) # Divide by 7 to get the average
  )
print(data_w)
```

### Verify the complete results of tranformed data

Before converting the file into a CSV file, I organized the data by sorting the data by total points and after that pre_ratings in descending order. 
```{r}
complete_results <- data_w %>% 
  select(Name, State, Points, pre_ratings, average_opponent_pre_rating) %>%
  arrange(desc(Points), desc(pre_ratings))
```

### Convert dataset into a CSV file
I converted the complete_results data into a CSV file.
```{r}
write.csv(complete_results, file='tournamentinfo_updatedsummary.csv', row.names = FALSE, quote = FALSE)
```

### View the results
```{r}
# Read the text file
results_data <- read_lines("https://raw.githubusercontent.com/pujaroy280/DATA607Project1/main/tournamentinfo_updatedsummary.csv")
print(results_data)
```

### Conclusion

Throughout this project, I learned how to load, extract, clean and transform and convert data into a CSV file.